### 2020-02-18 thesis/paper.tex
edit

\part{Testing}
\chapter{Experiments}\label{cha:experiments}
To show the value of the new tool different scenarios created. The goals of the tool are to improve on the ease of finding complex cluster structures and improve on the result in comparison to only using simple clustering methods. Another aim is to visualize the data and clustering results such that an expert user can make a well informed decision on which solutions best fit the data.
\section{Solutions better than Base-Clusterings}
\subsubsection{k-Means}
Creating base clusterings and applying consensus functions are well known methods, but there was previously no tool allowing to perform these steps as easily with a visual interface. Using this tool, a user is able to find better results than the individual ones created by sampling the parameters for the different clustering algorithms. This is the case as the combined knowledge through consensus clustering can be more powerful than an individual algorithm.

@@

Now as this data-set is quite simple let us create the base clusterings using Lloyd's k-Means algorithm. To increase the knowledge that can be extracted from the results, $k$ can simply be sampled in a large range. In this example, it is sampled in the range of $2$ to $20$ computing $6$ results for each $k$, see Figure \ref{fig:better_base_wf}. (Seed=$1$) Using such a larger range, even if an approximate $k$ is known is often useful as the additional information of possible smaller local clusters can be integrated. Also, as a poor value for $k$ leads to less robust results, the clusterings should differ more strongly and the randomness of the result should not make the information from small clusters to dominant.

@@

With the information of all clusterings taken into account through consensus clustering, the consensus result has an NMI of $1$ when compared with the ground truth, showing that this combined result is better than any individual run. Still it is important to note that the Lifetime criterion may fail to find an ideal choice for $k$ automatically. To overcome this, DICLENS can alternatively be used as it is better at defining $k$ although it has a much higher computational cost. It is also always possible to define $k$ by hand whenever this information is available.

@@

\section{Unobtainable Solutions}
Single clustering algorithms may not be able to find the structure of a data-set when different kinds of structures are present for different clusters. Combining results of different clustering algorithms may solve this problem, as correctly identified structures should be robust throughout multiple runs of a fitting algorithm while they should not be robust whenever the algorithm doesn't fit the structure. To illustrate this a data-set was created combining a density based cluster with Gaussian clusters with different deviations, see Figure \ref{fig:unob}.

@@

To produce base clusterings k-Means was run choosing $k$between $2$ and $20$ creating 5 results each and DBSCAN was run with $minPTS=5$ and $\epsilon$ between $0.01$ and $0.5$ taking 100 samples. For reproducability the seed was fixed at $1$ and trivial solutions were omitted. Also, these ranges were chosen once again such that a large diversity of solutions is obtained while not creating trivial clusterings. The workflow can be seen in Figure \ref{fig:unob_wf}.

@@

To solve this consensus clustering can be performed. As discussed previously both kinds of results have produced robust clusterings and the most relevant information can be combined to a better common result. In this example DICLENS was selected, fixing $k=4$ and the final result was computed. As can be seen in Figure \ref{fig:unob_consensus} using the information available from all solutions, both k-Means and DBSCAN, finds a result very close to the ground truth. This result has only few misclassified points, most noticeably the two red points outside the u-shaped cluster. The NMI of the consensus result is $0.978$ while the best result (Figure \ref{fig:unob_best}) has an NMI of $0.872$. This large improvement shows the power of using consensus clustering when different structures are present in the data.

@@

Visually, it is already difficult to decide how many clusters can be seen. The points on the left and right can be seen as one or two clusters each, but one could even claim that one side consists of one and the other two clusters. For this small data-set such claims can still easily be made but with more dimensions such thoughts become more complex. For this reason, I will show how different results can be found using this tool, by regarding the robustness of the results.

To generate base clusterings k-Means was used, varying the value for $k$ between $2$ and $10$ and computing $10$ samples for each value. (Seed=1) Trivial solutions were omitted and no ground truth added as none is available. The heat map for the resulting clusterings can be seen in Figure \ref{fig:multi_heatmap}

@@

In this heat map hierarchical structures of similar clustering results can already be observed. This structure can more clearly be defined in the OPTICS reachability plot, as seen in Figure \ref{fig:multi_optics} whereby the threshold is set such that valleys are separated and noise is left out. This way meta-clusters are defined which each represent a group of similar and robust clustering solutions.

@@

To create result candidates the consensus is computed with the Co-Association Matrix algorithm with Average Link strategy (Lifetime criterion) and letting the function guess $k$ itself. Thereby, as many consensus results are obtained as meta-clusters are available each representing the merged knowledge of the meta-cluster. The most interesting results should then be the ones which included many base clusterings, i.e. where the meta-cluster was large. This should be a fair assumption as the definition of the threshold fixes how similar clusterings need to be to be deemed robust and meta-clusters with many solutions should therefor be especially robust. For this reason, the number of merged base clusterings can be used as an implicit ranking of the consensus solutions. Figure \ref{fig:multi_consres} shows the consensus results for the four largest meta-clusters.

@@

Analyzing those results, one can see that the intuition mentioned before was also found in the consensus results for the largest meta-clusters. The largest meta-cluster, which included 19 base clusterings, resulted in a consensus clustering whereby $6$ clusters were found (see Figure \ref{fig:multi_c6}). This result can be seen as the one ranked highest and most robust as it includes the largest number of clusterings deemed robust. The second and forth largest meta-clusters found that the left and right group should be in a common cluster respectively (Figure \ref{fig:multi_c5_2} and \ref{fig:multi_c5_1}) while the third largest one considers both of them as one large cluster each (Figure \ref{fig:multi_c4}).

Finding these results can be helpful on top of only creating one final result, as knowledge may be gained. In this simple case it is easy to recognize how the left and right groups of points relate to each other but for more complex data-sets this may difficult to see. Here, it would for example be beneficial to look at what the underlying reason for the layout of the points may be, such that one can properly reason why a group of points should be one or two clusters. Also, note that in no part of this process quality metrics are needed, only the definition of the meta-cluster through the threshold should be done in a sensible way. In general, a good choice for this should be to visually analyze which cut-off value creates visually pleasing valleys, as well as considering how fine-grained one wants the differences between the results to be.

This way of finding multiple solutions may also prove useful when a data-set may be clustered with different goals. For example, clustering patient data could be done to find groups patients with different kinds of cancer or different kinds of diabetes. Even though both may cluster on the same data-set, the desired clusterings are likely to differ and different algorithms and parameter sets may reflect this.

@@

\section{Improvements for the Tool}
To tool created already proves useful for finding good clustering solutions, still further work should be done such that it becomes more usable.

A downside of the current implementation is that the focus of it lies solely on finding the best clustering result. This brings the problem that the methods implemented may not be fast enough when using a large data-set. To improve on this, the next steps should be to further optimize the methods for parallel environments and possibly implement similarly accurate but overall faster methods. This is especially true for the consensus functions as they can be very costly depending on the number of clusters in the individual results, target cluster number, points and dimensions in the data. Different functions runtime depends on different subsets of these values and also vary in their accuracy. Here more work should be put into finding the generally most accurate and performant method.

@@

\section{Research Consensus Clustering}
There has been much research on the topic of consensus clustering, still some aspects have not been analyzed thoroughly enough. While most papers on consensus functions show how their method is better than previous methods, the each pick different data-sets and become incomparable to a researcher. As most of them are difficult to implement and only few publicly share an implementation, testing the methods is also a difficult task. Still, as those methods are promising creating a large scale comparison of a variety of consensus functions regarding runtime, accuracy and requirements would be beneficial.

>>>>

\part{Testing}
\chapter{Experiments}\label{cha:experiments}
To show the value of the new tool different scenarios created. The goals of the tool are to improve on the ease of finding complex cluster structures and improve on the result in comparison to only using simple clustering methods. Another aim is to visualize the data and clustering results such that an expert user can make a well informed decision on which solutions best fit the data. For all experiments performed the Variation of Information distance measure was used, as it produced better results and the random seed was fixed at $1$ for reproducability of the experiments. The parameters for OPTICS were also left at their default values.

\section{Solutions better than Base-Clusterings}
\subsubsection{k-Means}
Creating base clusterings and applying consensus functions are well known methods, but there was previously no tool allowing to perform these steps as easily with a visual interface. Using this tool, a user is able to find better results than the individual ones created by sampling the parameters for the different clustering algorithms. This is the case as the combined knowledge through consensus clustering can be more powerful than an individual algorithm.

@@

Now as this data-set is quite simple let us create the base clusterings using Lloyd's k-Means algorithm. To increase the knowledge that can be extracted from the results, $k$ can simply be sampled in a large range. In this example, it is sampled in the range of $2$ to $20$ computing $6$ results for each $k$, see Figure \ref{fig:better_base_wf}. Using such a larger range, even if an approximate $k$ is known is often useful as the additional information of possible smaller local clusters can be integrated. Also, as a poor value for $k$ leads to less robust results, the clusterings should differ more strongly and the randomness of the result should not make the information from small clusters to dominant.

@@

With the information of all clusterings taken into account through consensus clustering, the consensus result has an NMI of $1$ when compared with the ground truth, showing that this combined result is better than any individual run, as it is not only similar but equal to the ground truth. Still it is important to note that the Lifetime criterion may fail to find an ideal choice for $k$ automatically. To overcome this, DICLENS can alternatively be used as it is better at defining $k$ although it has a much higher computational cost. It is also always possible to define $k$ by hand whenever this information is available.

@@

\section{Unobtainable Solutions}
Single clustering algorithms may not be able to find the structure of a data-set when multiple different kinds of structures are present for individual clusters. Combining results of different clustering algorithms may solve this problem, as correctly identified structures should be robust throughout multiple runs of a fitting algorithm while they should not be robust whenever the algorithm doesn't fit the structure. To illustrate this a data-set was created combining a U-shaped density based cluster with Gaussian clusters with different deviations, see Figure \ref{fig:unob}.

@@

To produce base clusterings k-Means was run choosing $k$ between $2$ and $20$ creating 5 results each and DBSCAN was run with $minPTS=5$ and $\epsilon$ between $0.01$ and $0.5$ taking 100 samples. The seed was again fixed at $1$  for reproducability and trivial solutions were omitted. Also, these ranges were selected such that a large diversity of solutions is obtained while not creating trivial clusterings. Trivial clusterings are not desired in this process as all points either being in one or each point in an individual cluster is no useful information. The workflow can be seen in Figure \ref{fig:unob_wf}.

@@

To solve this consensus clustering can be performed. As discussed previously both kinds of results have produced robust clusterings and the most relevant information can be combined to a better common result. In this example DICLENS was selected, fixing $k=4$ and the final result was computed. As can be seen in Figure \ref{fig:unob_consensus}, using the information available from all solutions, both k-Means and DBSCAN runs, finds a result very close to the ground truth. This result has only few misclassified points, most noticeably the two red points outside the U-shaped cluster. The NMI of the consensus result is $0.978$ while the best result of a single run (Figure \ref{fig:unob_best}) has an NMI of $0.872$. This large improvement shows the power of using consensus clustering when different structures are present in the data.

@@

Visually, it is already difficult to decide how many clusters can be seen. The points on the left and right can be seen as one or two clusters each, but one could even claim that one side consists of one and the other two clusters. For this small data-set such claims can still easily be made but with more dimensions such thoughts become more complex. For this reason, I will show how different results can be found using this tool, by regarding indicators of robustness for the results.

To generate base clusterings k-Means was used, varying the value for $k$ between $2$ and $10$ and computing $10$ samples for each value. Trivial solutions were omitted and no ground truth was added as none is available. With the workflow defined, it is executed also performing OPTICS for meta-clustering using the default MinPTS and $\epsilon$. With the ordering resulting from the OPTICS algorithm, the heat map for the resulting clusterings is created and can be seen in Figure \ref{fig:multi_heatmap}.

@@

In this heat map hierarchical structures of similar clustering results can already be observed. Darker rectangles near the diagonal show regions with similar clustering results, indicating their robustness as many different runs found similar results. This meta-cluster structure can more clearly be defined in the OPTICS reachability plot, as seen in Figure \ref{fig:multi_optics}, whereby the threshold is set such that valleys are separated and noise is left out. With this, the meta-clusters are defined each representing a group of similar and therefor hopefully robust clustering solutions.

@@

To create result candidates, the consensus is computed with the Co-Association Matrix algorithm with Average Link strategy (Lifetime criterion) and letting the function guess $k$ by itself. This way, as many consensus results are obtained as meta-clusters are available, each one representing the merged knowledge of its meta-cluster. The most interesting results should then be the ones which included many base clusterings, i.e. where the meta-cluster was large. This should be a fair assumption as the definition of the threshold fixes how similar clusterings need to be to be deemed robust and meta-clusters with many solutions should therefor be especially robust. For this reason, the number of merged base clusterings can be used as an implicit ranking of the consensus solutions. Figure \ref{fig:multi_consres} shows the consensus results for the four largest meta-clusters with the corresponding number of included base clusterings.

@@

Analyzing those results, one can see that the intuition mentioned before was also found by the consensus results for the largest meta-clusters. The largest meta-cluster, which included 19 base clusterings, resulted in a consensus clustering whereby $6$ clusters were found (see Figure \ref{fig:multi_c6}). This result can be seen as the highest ranked one and therefor most robust as it includes the largest number of clusterings deemed robust. The second and forth largest meta-clusters found that the left or right group should be in a common cluster respectively (Figure \ref{fig:multi_c5_2} and \ref{fig:multi_c5_1}) while the third largest one considers both of them as one large cluster each (Figure \ref{fig:multi_c4}). The robustness of the result with $6$ clusters can also further be seen when the consensus over all base clusterings is computed. Even without setting a parameter $k$ for this computation the result is equivalent to the one in Figure \ref{fig:multi_c6}.

Finding more than one possible result can be helpful on top of only creating one final one, as knowledge may be gained. In this simple case it is easy to recognize how the left and right groups of points relate to each other but for more complex data-sets this may difficult to see. Here, it would for example be beneficial to look at what the underlying reason for the layout of the points may be, such that one can properly reason why a group of points should be one or two clusters. Also, note that in no part of this process quality metrics are needed, only the definition of the meta-cluster through the threshold should be done in a sensible way. In general, a good choice for this should be to visually analyze which cut-off value creates visually pleasing valleys, as well as considering how fine-grained one wants the differences between the results to be.

This way of finding multiple solutions may also prove useful when a data-set may be clustered with different goals. For example, clustering patient data could be done to find groups of patients with different kinds of cancer or different kinds of diabetes. Even though both may cluster on the same data-set (for example vectorized patient data), the desired clusterings are likely to differ and different algorithms and parameter-sets may reflect this.

@@

\section{Improvements for the Tool}
Even though the created tool already proves useful for finding good clustering solutions, further work should still be done such that it becomes more generally usable.

A downside of the current implementation is that the focus of it lies solely on finding the best clustering result. This brings the problem that the methods implemented may not be fast enough when using a large data-set. To improve on this, the next steps should be to further optimize the methods for parallel environments and possibly implement similarly accurate but overall faster methods. This is especially true for the consensus functions as they can be very costly depending on the number of clusters in the individual results, target cluster number, points and dimensions in the data. Different functions runtime depend on different subsets of these values and also vary in their accuracy. Here more work should be put into finding the generally most accurate and performant method.

@@

\section{Research Consensus Clustering}
There has been much research on the topic of consensus clustering, still some aspects have not been analyzed thoroughly enough. While most papers on consensus functions show how their method is better than previous methods, they each pick different data-sets and become incomparable to a researcher. As most of them are difficult to implement and only few authors publicly share an implementation, testing the methods is also a difficult task. Still, as those methods are promising, creating a large scale comparison of a variety of consensus functions regarding runtime, accuracy and requirements/assumptions would be beneficial.
